### 2018-04-24 source/intro/RIME.tex
proofread intro

\chapter{Introduction to Mathematical Framework: the RIME Formalism}
\pg
In this section, we introduce the mathematical framework which forms the basis for our algorithmic work.
\label{section.RIME}
While radio interferometry has historically been a complex business (\citepads{2001isra.book.....T}), modern radio interferometry is based on the Radio Interferometer's Measurement Equation, a powerful and elegant formalism which underpins modern calibration and imaging algorithms. In this section, I will give a simplified account of this formalism as described in \citepads{2011A&A...527A.106S} and its companion papers\footnote{\citepads{2011A&A...527A.107S}, \citepads{2011A&A...527A.108S} and \citepads{2011A&A...531A.159S}}. This will form the theoretical basis on which further sections will expand. I cannot overstate the importance of the RIME as the foundation of modern calibration and imaging algorithms.

\pg
For a mathematically rigorous description of the RIME, the references used in this work (particularly \citepads{2011A&A...527A.108S} and \citepads{2011A&A...531A.159S}) are obviously the first place to look. We frame the RIME in the continuity of previous theoretical frameworks of radio interferometry, while also describing it in the context of contemporary software packages and algorithmic tools.

@@

\pg
Consider a single quasi-monochromatic point source in the sky. Its signal at a point in time and space can be described by a complex vector $\evector$. We can then represent physical processes which affect this signal's propagation using \emph{Jones matrices}. In other words:
\begin{equation}
\volt = \Jones\evector
\end{equation}
where $\volt$ is the voltage measured by our antenna, $\Jones$ the Jones matrix describing the \emph{net} propagation effects which affect our source's signal. Note that $\Jones$ can be written as a series of matrix products, each individual matrix describing a physical phenomenon. They are then called a \emph{Jones chain}, and the resulting $\Jones$ is often called the \emph{total Jones matrix}.

@@

\pg
If we choose to use $xyz$ as our coordinates, with $z$ the direction of propagation of our signal, then [Hamaker and Bregman 1996] show that the following relation holds:
\begin{equation}
2 \evector \evector^\Herm = 2  \begin{bmatrix} <e_x e_x^\star> & <e_x e_y^\star> \\ <e_y e_x^\star> & <e_y e_y^\star> \end{bmatrix} = \begin{bmatrix} I+Q & U+iV \\ U-iV & I-Q \end{bmatrix} = \Bmatrix
\end{equation}

@@

\pg
We shall use the conventional coordinate system and notations (\citepads{2011A&A...527A.106S}, \citepads{2001isra.book.....T}, etc). We thus have a $z$ axis pointing towards the phase centre, and an antenna $p$ at coordinates $\U_p=(u_p,v_p,w_p)$. The phase difference between $p$ and phase centre (where, by definition, $\U=0$) for a signal arriving from direction $\pmb{\sigma}$ is then:
\begin{equation}
\kappa_p=2\pi i (( u_p l + v_p m + w_p (n-1) ))
\end{equation}

where $n=\sqrt{1-l^2-m^2}$. Here, $l,m,n$ are the direction cosines\footnote{The $n-1$ term in the exponential is due to the fact that, at phase centre, $\kappa_p=0$ for all $\U$ and $n=\sqrt{1-0-0}=1$.} of $\pmb{\sigma}$, and $\U$ is defined in units of wavelength\footnote{This means that, for a given baseline $pq$, $\U$ will vary as a function of frequency.}.

\pg 
Having done this, we can now introduce a scalar \emph{K-Jones} matrix to represent the effect of this phase delay. Since it is scalar, it will commute with all other Jones matrices [see smirnov 2011]. It can be be written as:
\begin{equation}
\Kjones_p = e^{-i \kappa_p}  = e^{-2\pi i ( u_p l + v_p m + w_p (n-1) )}
\end{equation}

@@

\pg
So far, $\Bmatrix$ represents the signal from a single point source. However, the formalism can be extended to multiple point sources fairly easily. This is because electric fields -- and therefore radio signals -- are additive. If our sky contains $s$ point sources, then our RIME becomes:
\begin{equation}
\Vis_{pq} = \sum_s \Jones_{sp} \Bmatrix_s \Jones_{sq}^\Herm
\end{equation}

@@

\pg
Here, effects such as our radio beam lie in the Jones matrices. By performing the analysis of [thompson 2001, sect 3.1], we can rewrite this expression in terms of a sine projection of the sphere onto the $(l,m)$ plane tangential at field centre. The integral then becomes:
\begin{equation}
\Vis_{pq} = \int_{l,m} \Jones_p(\direction) \Bmatrix(\direction) \Jones_q^H(\direction)\frac{dl dm}{n}
\end{equation}

@@

\pg
Earlier calibration models and algorithm rely on the hypothesis that all baselines see the same sky. This is the \emph{fundamental premise of self-calibration}. The topic of self-cal will be covered in \cref{section.calibration.2gc}. This premise only holds when all Direction-Dependent Effects (henceforth referred to as DDEs) are identical for all baselines (and therefore for all antennas). This is a necessary condition for the apparent sky $\Bmatrix_{pq}$ to be the same for all baselines\footnote{$\Bmatrix_{pq}$ would then correspond to the `true' sky attenuated by the power beam, in the traditional view}, as it allows us to write:
\begin{align}
\Ejones_p(\lvec) \Wjones_p(\lvec)&= \Ejones(\lvec) \Wjones(\lvec)\\
\Bmatrix_{pq}(\lvec) = \Bmatrix_{app}(\lvec) &= \Ejones(\lvec) \Wjones(\lvec) \Bmatrix)(\lvec) \Wjones^H(\lvec) \Ejones^H(\lvec)
\end{align}

@@

\pg
Until this point, we have not considered any kind of time variability. In effect, \cref{eq.fullsky.RIME} describes the snapshot taken by an interferometer for a single measurement. Limiting ourselves to this scenario, however, constrains the sampling of the $uv$-plane which can be performed. Indeed, interferometers generally rely on the Earth's rotation to rotate their baselines $pq$, thus increasing $uv$-coverage at no additional cost. For \cref{eq.fullsky.RIME} to hold throughout an observation, we must therefore assume that $B_{app}$ remains constant in time over the course of the observation. In effect, this means assuming that
\begin{equation}\label{eq.selfcal.requirement}
\Ejones_p(t,\lvec) = \Ejones_p(\lvec) = \Ejones(\lvec) \quad \forall (t,p)
\end{equation}

@@

\pg
Of course, our Jones matrices are not the only time-dependent part of the RIME formalism. To image with interferometric arrays, radio astronomers rely on the Earth's rotation to better sample $uv$-space, a technique referred to as \emph{super-synthesis} (\cite{supersynthesis}\footnote{\href{http://www.jpier.org/PIERB/pierb22/05.10032105.pdf}{(link to paper)}}). This means that $\uvec$ becomes a \emph{time-dependent} variable.

@@

\pg
In practice, measurements made by any interferometric array is actually averaged over the integration time of the measurement, as well as the frequency channel size. A better formulation is therefore to explicitly write the RIME as an integration over time and frequency bins:
\begin{align}
\langle \Vis_{pw} \rangle &= \frac{1}{\Delta t \Delta \nu} \int_{t_0}^{t_1} \int_{\nu_0}^{\nu_1} \Vis_{pq}(t,\nu) dv dt\\
                          &= \frac{1}{\Delta t \Delta \nu} \int_{t_0}^{t_1} \int_{\nu_0}^{\nu_1} \Jones_p(t,\nu) \Bmatrix \Jones_q^H(t,\nu)
\end{align}

@@

\pg Smearing increases with baseline length ($\uvec_{pq}$) and distance from phase centre ($l,m$). The noise amplitude, however, does not decrease. This means that smearing results in a decrease in sensitivity, which becomes significant when attempting to do wide-field, high-resolution images (see section [PLUG GROTH WORK SHAMELESSLY]).

@@

\pg
This noise sets a hard limit on the sensitivity achieved for a given observation, and `reaching the noise' becomes the gold standard of calibration (\citepads{2011A&A...527A.107S}). One must then take $\sigma_{pq}$ into account properly when solving for the Jones matrices.

>>>>

\chapter{Mathematical Framework: the RIME Formalism}
\pg
In this section, we introduce the mathematical framework which forms the basis for our algorithmic work.
\label{section.RIME}
While radio interferometry has historically been a complex business \citepads{2001isra.book.....T}, modern radio interferometry is based on the Radio Interferometer's Measurement Equation, a powerful and elegant formalism which underpins modern calibration and imaging algorithms. In this section, I will give a simplified account of this formalism as described in \citetads{2011A&A...527A.106S} and its companion papers\footnote{\citetads{2011A&A...527A.107S}, \citetads{2011A&A...527A.108S} and \citetads{2011A&A...531A.159S}}. This will form the theoretical basis on which further sections will expand. I cannot overstate the importance of the RIME as the foundation of modern calibration and imaging algorithms.

\pg
For a mathematically rigorous description of the RIME, the references used in this work (particularly \citetads{2011A&A...527A.108S} and \citetads{2011A&A...531A.159S}) are obviously the first place to look. We frame the RIME in the continuity of previous theoretical frameworks of radio interferometry, while also describing it in the context of contemporary software packages and algorithmic tools.

@@

\pg
Consider a single quasi-monochromatic point source in the sky. Its signal at a point in time and space can be described by a complex vector $\evector$. We can then represent physical processes which affect this signal's propagation using \emph{Jones matrices}. In other words:
\begin{equation}
\volt = \Jones\evector
\end{equation}
where $\volt$ is the voltage measured by our antenna, and $\Jones$ the Jones matrix describing the \emph{net} propagation effects which affect our source's signal. $\Jones$ can be written as a series of matrix products, each individual matrix describing a physical phenomenon. They are then called a \emph{Jones chain}, and the resulting $\Jones$ is often called the \emph{total Jones matrix}.

@@

\pg
If we choose to use $xyz$ as our coordinates, with $z$ the direction of propagation of our signal, then \citetads{1996A&AS..117..137H} show that the following relation holds:
\begin{equation}
2 \evector \evector^\Herm = 2  \begin{bmatrix} <e_x e_x^\star> & <e_x e_y^\star> \\ <e_y e_x^\star> & <e_y e_y^\star> \end{bmatrix} = \begin{bmatrix} I+Q & U+iV \\ U-iV & I-Q \end{bmatrix} = \Bmatrix
\end{equation}

@@

\pg
We shall use the conventional coordinate system and notations (\citetads{2011A&A...527A.106S}, \citetads{2001isra.book.....T}, etc). We thus have a $z$ axis pointing towards the phase centre, and an antenna $p$ at coordinates $\U_p=(u_p,v_p,w_p)$. The phase difference between $p$ and phase centre (where, by definition, $\U=0$) for a signal arriving from direction $\pmb{\sigma}$ is then:
\begin{equation}
k_p=2\pi i (( u_p l + v_p m + w_p (n-1) ))
\end{equation}

where $n=\sqrt{1-l^2-m^2}$. Here, $l,m,n$ are the direction cosines\footnote{The $n-1$ term in the exponential is due to the fact that, at phase centre, $k_p=0$ for all $\U$ and $n=\sqrt{1-0-0}=1$.} of $\pmb{\sigma}$, and $\U$ is defined in units of wavelength\footnote{This means that, for a given baseline $pq$, $\U$ will vary as a function of frequency.}.

\pg 
Having done this, we can now introduce a scalar \emph{K-Jones} matrix to represent the effect of this phase delay. Since it is scalar, it will commute with all other Jones matrices (see \citetads{2011A&A...527A.106S}). It can be be written as:
\begin{equation}
\Kjones_p = e^{-i k_p}  = e^{-2\pi i ( u_p l + v_p m + w_p (n-1) )}
\end{equation}

@@

\pg
So far, $\Bmatrix$ represents the signal from a single point source. However, the formalism can be extended to multiple point sources. This is because electric fields -- and therefore radio signals -- are additive. If our sky contains $s$ point sources, then our RIME becomes:
\begin{equation}
\Vis_{pq} = \sum_s \Jones_{sp} \Bmatrix_s \Jones_{sq}^\Herm
\end{equation}

@@

\pg
Here, effects such as our radio beam lie in the Jones matrices. By performing the analysis shown in Sect. 3.1 of \citetads{2001isra.book.....T}, we can rewrite this expression in terms of a sine projection of the sphere onto the $(l,m)$ plane tangential at field centre. The integral then becomes:
\begin{equation}
\Vis_{pq} = \int_{l,m} \Jones_p(\direction) \Bmatrix(\direction) \Jones_q^H(\direction)\frac{dl dm}{n}
\end{equation}

@@

\pg
Earlier calibration models and algorithm rely on the hypothesis that all baselines see the same sky. This is the \emph{fundamental premise of self-calibration}. The topic of self-cal is covered in \cref{section.calibration.2gc}. This premise only holds when all Direction-Dependent Effects (henceforth referred to as DDEs) are identical for all baselines (and therefore for all antennas). This is a necessary condition for the apparent sky $\Bmatrix_{pq}$ to be the same for all baselines\footnote{$\Bmatrix_{pq}$ would then correspond to the `true' sky attenuated by the power beam, in the traditional view}, as it allows us to write:
\begin{align}
\Ejones_p(\lvec) \Wjones_p(\lvec)&= \Ejones(\lvec) \Wjones(\lvec)\\
\Bmatrix_{pq}(\lvec) = \Bmatrix_{app}(\lvec) &= \Ejones(\lvec) \Wjones(\lvec) \Bmatrix)(\lvec) \Wjones^H(\lvec) \Ejones^H(\lvec)
\end{align}

@@

\pg
Until this point, we have not considered any kind of sky variability in time. In effect, \cref{eq.fullsky.RIME} describes the snapshot taken by an interferometer for a single measurement. Limiting ourselves to this scenario, however, strongly limits the sampling of the $uv$-plane which can be performed. Indeed, interferometers generally rely on the Earth's rotation to rotate their baselines $pq$, thus increasing $uv$-coverage at no additional cost.This technique is known as \emph{super-synthesis} (\cite{supersynthesis}\footnote{\href{http://www.jpier.org/PIERB/pierb22/05.10032105.pdf}{(link to paper)}}) For \cref{eq.fullsky.RIME} to hold throughout an observation, we must therefore assume that $B_{app}$ remains constant in time over the course of the observation. In effect, this means assuming that
\begin{equation}\label{eq.selfcal.requirement}
\Ejones_p(t,\lvec) = \Ejones_p(\lvec) = \Ejones(\lvec) \quad \forall (t,p)
\end{equation}

@@

\pg
Of course, our Jones matrices are not the only time-dependent element of the RIME formalism. As mentioned above, to image with interferometric arrays, radio astronomers rely on the Earth's rotation to better sample $uv$-space. This means that $\uvec$ becomes a \emph{time-dependent} variable.

@@

\pg
In practice, measurements made by any interferometric array are actually averaged over the integration time of the measurement, as well as the frequency channel size. A better formulation is therefore to explicitly write the RIME as an integration over time and frequency bins:
\begin{align}
\langle \Vis_{pw} \rangle &= \frac{1}{\Delta t \Delta \nu} \int_{t_0}^{t_1} \int_{\nu_0}^{\nu_1} \Vis_{pq}(t,\nu) dv dt\\
                          &= \frac{1}{\Delta t \Delta \nu} \int_{t_0}^{t_1} \int_{\nu_0}^{\nu_1} \Jones_p(t,\nu) \Bmatrix \Jones_q^H(t,\nu)
\end{align}

@@

\pg Smearing increases with baseline length ($\uvec_{pq}$) and distance from phase centre ($l,m$). The noise amplitude, however, does not decrease. This means that smearing results in a decrease in sensitivity, which becomes significant when attempting to do wide-field, high-resolution images (see section \textcolor{red}{[PLUG GROTH WORK SHAMELESSLY]}).

@@

\pg
This noise sets a hard limit on the sensitivity achieved for a given observation, and `reaching the noise' becomes the gold standard of calibration \citepads{2011A&A...527A.107S}. One must then take $\sigma_{pq}$ into account properly when solving for the Jones matrices.
